---
title: "The Educational Divide: Understanding the Interaction Between Educational Attainment and Racial Identity in American Voting Patterns"
subtitle: "A Multinomial Analysis of the Cooperative Election Study Common Content Dataset"
author: 
  - Jinyan Wei
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "This study examines the impact of the Dobbs decision and subsequent abortion bans on infant mortality rates in the United States, focusing on data from 2021 to 2022. Using a Difference-in-Differences (DID) approach, we analyze trends in states where abortion remained legal compared to those where it became illegal after June 2022. Our results indicate that states enforcing abortion bans experienced an increase of 0.285 infant deaths per 1,000 live births, with compounded effects observed during the post-injunction period. Maternal age and race also emerged as significant predictors, with Black mothers facing disproportionately higher infant mortality rates. These findings underscore the compounded public health implications of restrictive abortion policies and the urgent need for targeted interventions to address racial and demographic disparities. By utilizing recent, high-quality data and rigorous modeling, this study provides critical insights into the intersection of reproductive rights and infant health outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 90
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
library(knitr)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(nnet)
```

```{r}
#| include: false
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)

# Load the data
analysis_data<-read_parquet(here::here("data", "02-analysis_data", "cleaned_demographic_data.parquet"))

```

# Introduction

The relationship between educational attainment and voting behavior has long been a
subject of scholarly interest, particularly as the American electorate becomes
increasingly diverse and polarized. Understanding how education levels interact with
racial identity to shape political preferences is crucial for comprehending contemporary
voting patterns and their implications for democratic participation.

This study examines the complex interplay between educational attainment and racial
demographics in determining voting choices, using comprehensive data from the 2022
Cooperative Election Study (CES). The CES dataset, comprising 60,000 respondents through
matched random sampling by YouGov, provides detailed insights into voter preferences,
demographic characteristics, and political behavior across all U.S. states. Our analysis
addresses a critical gap in the literature by investigating how the impact of education on
voting decisions varies across racial groups, accounting for socioeconomic differences and
access to opportunities.

Using a multinomial logistic regression framework, we analyze how educational attainment
interacts with racial identity to influence voting patterns, while controlling for key
demographic and socioeconomic factors including gender, urbanicity, gun ownership, and
education loan status. The model incorporates both direct effects and interaction terms to
capture the nuanced relationships between these variables and voting preferences. This
methodological approach allows us to estimate the differential impacts of education across
racial groups while accounting for other important determinants of political behavior.

The findings indicate that the relationship between educational attainment and voting
choice varies significantly across racial groups, suggesting that the political
implications of education are not uniform across demographic categories. This research
contributes to our understanding of contemporary American electoral behavior by providing
current evidence on how educational experiences and racial identity jointly shape
political preferences. These insights have important implications for understanding voter
behavior, demographic voting patterns, and the broader dynamics of political participation
in an increasingly diverse electorate.

The remainder of this paper is organized as follows: @sec-data details our data sources
and variable measurements, @sec-model presents our multinomial regression methodology and
model specifications, @sec-results discusses our empirical findings, and @sec-discussion
concludes with implications and directions for future research. Additional methodological
details and robustness checks are provided in [Appendix- @sec-data-details], [Appendix-
@sec-model-details], and [Appendix- @sec-survey].

# Data {#sec-data}

## Overview

We conduct our voting pattern analysis using the R programming language [@citeR],
leveraging data from the 2022 Cooperative Election Study [@CESdata]. The CES dataset,
constructed through matched random sampling by YouGov, offers a comprehensive overview of
voter preferences, demographic details, and political behavior in the United States. The
survey, designed to ensure representation at both state and national levels, incorporates
pre-election and post-election responses from 60,000 participants. This robust dataset
enables the analysis of voting behavior, party identification, policy preferences, and
their interactions with demographic variables.

In this study, we utilized several R packages to enhance data manipulation, modeling, and
visualization capabilities. The `tidyverse` package [@tidyverse] provided a suite of tools
for efficient data wrangling and exploration, while `arrow` [@arrow] facilitated the
management of large parquet files, ensuring seamless data input and output. The `janitor`
package [@janitor] supported data cleaning by offering tools to identify and rectify
quality issues within the dataset, and `lubridate` [@lubridate] simplified the handling of
date-time variables. To streamline file organization and improve reproducibility, we
relied on the `here` package [@here].

For modeling voting patterns, the `nnet` package [@nnet] was used to fit multinomial
logistic regression models, capturing the influence of demographic factors and their
interactions on vote choice. Results were summarized and visualized using the
`modelsummary` package [@modelsummary] and `ggplot2` [@ggplot2], respectively, ensuring
clarity and interpretability. Additionally, we employed `knitr` [@knitr] and `kableExtra`
[@kableExtra] to generate clean, reproducible tables and reports. Together, these packages
supported an efficient and transparent workflow for analyzing the CES dataset, enabling a
deeper understanding of the complex relationship between race, education, and voting
behavior.

## Measurement

The process of translating survey responses into a structured dataset for the CES 2022
analysis requires a systematic approach to measurement and data gathering. In this
research, we aim to investigate the factors influencing voter preferences, particularly
the interaction between education and race in shaping voting behavior. The CES dataset
captures responses to questions about voter choice, political attitudes, and demographic
characteristics, enabling a comprehensive analysis of election dynamics in the United
States. Survey items are carefully designed to address diverse aspects of voter behavior,
such as political affiliation, policy preferences, and party identification.

To ensure a representative sample, the CES employs matched random sampling techniques
stratified by demographics and geography. Respondents are recruited using various methods,
including online panels and targeted outreach, to reflect the diversity of the U.S.
electorate. After responses are collected, the data undergoes rigorous cleaning and
validation procedures to address inconsistencies, rectify missing values, and ensure the
dataset's accuracy. For instance, variables like education and race are recategorized into
uniform groups to standardize comparisons. Weighting is applied to adjust for potential
sampling biases, accounting for demographic disparities in age, gender, and state
representation.

The cleaned and validated data is then stratified and aggregated to examine voting
patterns and trends across different subgroups. Statistical models, including multinomial
logistic regression, are employed to analyze how race and education interact to influence
voting preferences, highlighting nuances in voter behavior across demographic lines. This
structured methodology transforms individual survey responses into actionable insights,
providing a detailed understanding of how demographic factors shape electoral outcomes. By
combining robust survey design with advanced data analysis techniques, this study captures
the complex dynamics of contemporary U.S. elections.

## Variables

Our analysis focuses on the following variables, with a specific emphasis on `vote_choice`
as the dependent variable:

-   `vote_choice`: The voting preference of individuals, serving as the dependent variable
    in our analysis. The categories include:
    -   **Democrat**: Individuals who voted for the Democratic candidate.
    -   **Republican**: Individuals who voted for the Republican candidate.
    -   **Independent**: Individuals who voted for a third-party or independent candidate.
    -   **Did not vote**: Individuals who did not participate in voting.
-   `education`: The highest level of education completed by individuals. Categories
    include:
    -   **High school or less**: Individuals with a high school diploma or lower education
        level.
    -   **Some college**: Individuals with some college education but no degree.
    -   **College graduate**: Individuals who have completed a bachelor's degree.
    -   **Postgraduate**: Individuals with education beyond a bachelor's degree.
-   `race_group`: The racial or ethnic group individuals identify with. Categories
    include:
    -   **White**: Individuals identifying as White.
    -   **Black**: Individuals identifying as Black or African American.
    -   **Hispanic**: Individuals identifying as Hispanic or Latino.
    -   **Asian**: Individuals identifying as Asian.
    -   **Other**: Individuals identifying with any other racial or ethnic group.
-   `gender`: The self-reported gender of individuals, categorized as:
    -   **Male**
    -   **Female**
-   `urbancity`: The type of community individuals reside in, categorized as:
    -   **Urban**: Dense metropolitan areas.
    -   **Suburban**: Residential areas near urban centers.
    -   **Rural**: Sparsely populated regions or countryside.
-   `gunown`: Indicates whether an individual owns a gun, categorized as:
    -   **Yes**
    -   **No**
-   `edloan`: Indicates whether an individual has education loans, categorized as:
    -   **Yes**
    -   **No**
-   `age`: The age of the individual, recorded as a continuous variable.

```{r}
#| label: fig-age-distribution
#| fig-cap: >
#|   Distribution of Respondent Age
#| echo: false

ggplot(analysis_data, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "brown", color = "black") +
  labs(
    x = "Age (Years)", y = "Count",
    caption = "Data source: CES 2022."
  ) +theme_classic()
 
```

@fig-age-distribution shows the distribution of respondent ages, with most individuals
concentrated between the ages of 30 and 60. The distribution appears relatively symmetric,
with no extreme outliers, suggesting that the sample covers a broad range of adult ages.
This balance ensures that the analysis includes perspectives from various life stages.

```{r}
#| label: fig-vote-choice
#| fig-cap: >
#|   Distribution of Vote Choice: The distribution of respondents' voting choices 
#|   in the dataset highlights that the majority voted for either the Republican 
#|   or Democrat candidates, with a smaller proportion selecting other candidates.
#| echo: false

library(ggplot2)

# Scatter plot of vote choice by age, colored by race group
ggplot(analysis_data, aes(x = age, y = as.factor(vote_choice), color = race_group)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.7, size = 2) +
  scale_color_brewer(palette = "Set1", name = "Race Group") +
  theme_classic() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14)
  ) +
  labs(
    title = "Vote Choice by Age and Race Group",
    x = "Age",
    y = "Vote Choice",
    caption = "Data Source: Cleaned Demographic Data"
  )
```

@fig-vote-choice shows the distribution of respondents' voting choices across different
age groups, categorized by race. The visualization highlights three main categories of
vote choice: Republican, Democrat, and Other. Each point represents an individual, with
its position along the x-axis corresponding to their age and its color indicating their
race group. This graph illustrates how voting preferences may vary across demographics,
showing a significant spread for both Republican and Democrat voters across all ages,
while the "Other" category remains sparsely populated. Observing the clustering patterns
within race groups can provide insights into potential demographic trends influencing
voting behavior.

```{r}
#| label: fig-race-group
#| fig-cap: >
#|   Distribution of Respondent Age: The age distribution shows a peak 
#|   around middle-aged respondents, with a relatively symmetric spread.
#| echo: false
education_dist <- analysis_data %>%
  group_by(race_group, education) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(race_group) %>%
  mutate(proportion = count / sum(count),
         percent = round(proportion * 100))

ggplot(education_dist, aes(x = race_group, y = proportion, fill = education)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = sprintf("%d%%", percent)), 
            position = position_stack(vjust = 0.5),
            color = "black", size = 3) +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.2)) +
  scale_fill_manual(
    values = c("lightblue", "darkgreen", "orange", "royalblue", "gold"),
    name = "Education Level"
  ) +
  labs(
    title = "Highest Educational Attainment by Race/Ethnicity",
    x = "Race Group",
    y = "Proportion",
    caption = "Note: Values may not sum to 100 percent due to rounding."
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  )
```

@fig-race-group visualizes the distribution of educational attainment across racial and
ethnic groups, highlighting disparities that may influence political behavior. "College
graduate" is the largest educational category across most groups, with a particularly high
proportion among Middle Eastern (64%) and Other (63%) populations. Meanwhile, "High school
or less" is notably prevalent among Native American (33%) and Hispanic (20%) respondents,
suggesting barriers to higher education for these groups. The graph underscores how the
interaction between race and education shapes opportunities and outcomes, making it a
crucial factor for analyzing voting behavior. For instance, research like @enos2018
emphasizes that education's political effects are context-dependent, moderated by local
demographics and historical inequalities, further supporting the need to explore this
interaction in understanding voting trends.

# Model {#sec-model}

Our modeling approach aims to quantify the relationship between demographic, contextual,
and behavioral factors and individuals' voting choices. For this analysis, we employ a
multinomial logistic regression (MLR) model to predict the probability of selecting each
voting category (`republican`, `democrat`, or `other`) based on a set of predictors. The
model is implemented using the `multinom` function in R, which estimates the log-odds of
each outcome relative to a reference category.

In this analysis, we use predictors that capture a broad range of individual
characteristics and contextual factors influencing voting behavior. Specifically, we
include variables such as `gender`, representing the respondent's self-reported gender;
`education`, capturing the highest level of education attained; and `race_group`,
representing the respondent's racial or ethnic background. Additional predictors include
`urbancity`, reflecting the type of community (urban, suburban, or rural) where the
respondent resides; `gunown`, indicating firearm ownership; and `edloan`, indicating the
presence of education loans. Finally, we incorporate `age` as a continuous variable.

To account for interactions between education and race, an extended version of the model
includes an interaction term (`education:race_group`). This term allows us to assess how
the relationship between education and voting choice varies across racial groups, offering
a more nuanced understanding of these predictors' combined effects.

The multinomial logistic regression assumes that the relative log-odds of any two outcomes
are linearly related to the predictors, and the independence of irrelevant alternatives
(IIA) assumption holds. This allows us to model the factors influencing voting preferences
in a consistent and interpretable framework.

## Model Parameters

The model predicts the probability of voting for each category (`republican`, `democrat`,
or `other`) by including the following predictor variables:

-   **Gender (`gender`)**: Categorical variable indicating the respondent's self-reported
    gender (`male` or `female`).
-   **Education (`education`)**: The highest level of education attained, categorized as
    `high school or less`, `some college`, `college graduate`, or `postgraduate`.
-   **Race Group (`race_group`)**: Categorical variable representing the respondent's
    racial or ethnic background (`White`, `Black`, `Hispanic`, `Asian`, or `Other`).
-   **Urbancity (`urbancity`)**: The type of community where the respondent resides
    (`urban`, `suburban`, or `rural`).
-   **Gun Ownership (`gunown`)**: Binary variable indicating whether the respondent owns a
    firearm (`Yes` or `No`).
-   **Education Loan (`edloan`)**: Binary variable indicating whether the respondent has
    outstanding education loans (`Yes` or `No`).
-   **Age (`age`)**: Continuous variable representing the respondent's age.

In the second model, an **interaction term** between `education` and `race_group` is
included to assess the combined effect of these predictors on voting preferences:

-   **Interaction Term (`education:race_group`)**: Captures how the effect of education on
    voting choice varies across racial groups.

We include variables like `urbancity`, `gunown`, and `edloan` as controls to account for
contextual and behavioral factors that could influence voting behavior. By modeling these
relationships, we aim to provide a balanced and detailed assessment of how demographic and
behavioral factors shape voting preferences.Background details and diagnostics are
included in [Appendix -@sec-model-details].

## Model Set-Up {#sec-model-setup}

In our analysis, we employ two multinomial logistic regression models to investigate
voting patterns. The first model includes primary predictor variables such as `gender`,
`education`, `race group`, `urbanicity`, `gun ownership`, `education loan status`, and
`age`. These variables capture key demographic and socioeconomic factors influencing
voting choices. The second model extends the first by introducing an interaction term
between education and race group to account for potential joint effects of these variables
on voting preferences.

Define $y_i$ as the voting choice for individual $i$, categorized as "Republican,"
"Democrat," or "Other." The probability $P(y_i = k)$ of an individual $i$ choosing
category $k$ is modeled as:

$$
\begin{aligned}
y_i|\eta_{ik}, \sigma &\sim \text{Multinomial}(P(y_i = k))\\
P(y_i = k) &= \frac{\exp(\eta_{ik})}{\sum_{j=1}^K \exp(\eta_{ij})} \\[1em]
\text{First Model}: \eta_{ik} &= \beta_{0k} + \beta_{1k} \cdot \text{Gender}_i + 
\beta_{2k} \cdot \text{Education}_i + 
\beta_{3k} \cdot \text{Race Group}_i \\ 
&+ \beta_{4k} \cdot \text{Urbancity}_i + 
\beta_{5k} \cdot \text{Gun Ownership}_i + 
\beta_{6k} \cdot \text{Education Loan}_i + 
\beta_{7k} \cdot \text{Age}_i \\[1em]
\text{Second Model}: \eta_{ik} &= \beta_{0k} + \beta_{1k} \cdot \text{Gender}_i + 
\beta_{2k} \cdot \text{Education}_i + 
\beta_{3k} \cdot \text{Race Group}_i \\ 
&+ \beta_{4k} \cdot \text{Urbancity}_i + 
\beta_{5k} \cdot \text{Gun Ownership}_i + 
\beta_{6k} \cdot \text{Education Loan}_i + 
\beta_{7k} \cdot \text{Age}_i \\ 
&+ \beta_{8k} \cdot (\text{Education}_i \times \text{Race Group}_i) \\[1em]
\end{aligned}
$$

The model is executed in R using the `multinom` function from the **`nnet`** package,
which fits the multinomial logistic regression model.These models allow us to examine both
the independent and combined effects of education and race on voting patterns while
controlling for other demographic and socioeconomic factors.

### Model justification {#sec-model-justify}

Scholars in political science have established that demographic factors such as gender,
education, race, and urbanicity, along with contextual variables like firearm ownership
and education loans, play significant roles in shaping voting behavior. Furthermore, the
interaction between `education` and `race_group` is well-supported in the literature, as
studies have shown that the impact of education on voting choice often varies by race due
to differences in socioeconomic context and access to opportunities. Including this
interaction term allows us to better capture these nuanced relationships and improve the
interpretability of the model.

We fit a frequentist multinomial logistic regression model, which is appropriate for
categorical outcomes with more than two levels. Coefficients are estimated using maximum
likelihood estimation, offering interpretable results without relying on prior
assumptions. Unlike Bayesian methods, no priors are applied. Additionally, we do not apply
weights to the observations, assuming the dataset accurately reflects the underlying
population. Future analyses, however, could incorporate weighting schemes to address
demographic imbalances or over- or under-representation of specific voting categories.

This model framework is well-suited for understanding the effects of multiple predictors
and their interactions on voting choice. Each coefficient represents the change in the
log-odds of selecting a given voting category relative to the reference category
(`republican`). By including main effects and interaction terms, the model provides a
theoretically informed and methodologically sound approach to analyzing voting
preferences. Diagnostics such as multicollinearity checks and assessments of the
independence of irrelevant alternatives (IIA) assumption further ensure the reliability of
the results.

# Results {#sec-results}

## Model Results

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false
first_model <-
  readRDS(file = here::here("models/first_model.rds"))
second_model <-
  readRDS(file = here::here("models/second_model.rds"))
```

The results from the two Bayesian models are summarized in @tbl-modelresults. The
interaction term between after_injunction and abortion_illegal is positive and
statistically significant (0.111 in first model and 0.280 in second model), indicating a
change in infant mortality rates associated with the combination of abortion bans and the
post-injunction period.

The results from the multinomial logistic regression models are summarized in
@tbl-modelresults. The first model examines the relationship between demographic and
socioeconomic factors and vote choice, while the second model includes interaction terms
between education and race group to explore their combined effects.

For the main effects, gender shows significant associations with vote choice. Non-binary
individuals are more likely to vote Democrat compared to Republicans in both models, with
a coefficient of 0.895 in the first model. Education also demonstrates varying impacts;
individuals with postgraduate education are less likely to vote Democrat compared to those
with high school education in Model 1, but this trend changes when interacting with race
group in Model 2. For instance, Black voters with a high school education are more
inclined to vote Democrat, highlighting the nuanced influence of race and education on
voting behavior.

Race group plays a critical role in shaping vote choice. Hispanic voters are positively
associated with voting for "other" candidates, with a coefficient of 0.555 in the first
model, reflecting a potential preference for non-mainstream parties. Similarly, urbanicity
shows a strong correlation with vote choice; rural voters are less likely to vote
Democrat, while suburban and urban voters lean more toward Democrats.

The second model's interaction terms reveal significant insights into how education and
race group intersect. For example, the interaction between postgraduate education and
Black voters suggests a greater likelihood of voting Democrat, while Hispanic voters with
high school education are more inclined to vote for other candidates.

The models performed well, with the residual deviance decreasing from 1825.802 in the
first model to 1795.585 in the second, indicating improved fit with the inclusion of
interaction terms. The AIC values further confirm this trend, decreasing from 1897.802 to
1931.585. These results underscore the complex interplay of demographic and socioeconomic
factors in influencing vote choice, while highlighting the importance of considering
intersectional effects for a more nuanced understanding.

# Discussion {#sec-discussion}

## Key Findings and Implications

The analysis reveals notable patterns in infant mortality rates influenced by policy and
demographic factors. For states where abortion remained legal, the post-injunction period
(after_injunction1) is associated with reductions in infant mortality rates, with a
decrease of 0.525 deaths per 1,000 live births in the first model and 0.947 in the second
model, compared to the pre-injunction period. In contrast, states where abortion became
illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with
increases of 3.365 deaths in the first model and 1.401 in the second model, relative to
states where abortion remained legal.

The Difference-in-Differences (DID) interaction term (after_injunction1 ×
abortion_illegal1) indicates that abortion bans during the post-injunction period resulted
in an additional increase of 0.111 deaths per 1,000 live births in the first model and
0.280 in the second model. These effects highlight the compounded impact of restrictive
abortion policies on infant health outcomes.

Demographic factors also play a significant role. Compared to mothers aged 15–19 years
(reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are
associated with progressively lower infant mortality rates, with reductions of 3.571,
5.732, and 6.155 deaths per 1,000 live births in the second model, respectively. Regarding
maternal race, Black or African American mothers have a significantly higher infant
mortality rate, with an increase of 6.604 deaths per 1,000 live births compared to Asian
mothers (reference level). In contrast, White mothers exhibit a lower infant mortality
rate, with a decrease of 2.562 deaths per 1,000 live births compared to Asian mothers.

These findings underscore the interplay of policy and demographic factors, emphasizing the
disproportionate burden of restrictive abortion policies and systemic inequities on infant
mortality rates.

## Racial and Socioeconomic Disparities

The results reveal stark racial disparities in infant mortality rates, with Black mothers
experiencing significantly higher rates compared to other racial groups. These disparities
are exacerbated by restrictive abortion policies, which disproportionately affect already
vulnerable populations. The intersection of systemic healthcare inequities, socioeconomic
challenges, and restrictive legislation creates compounded risks for marginalized
communities, leading to worse health outcomes. Addressing these inequities requires
targeted policies that ensure equitable access to maternal and infant healthcare,
particularly in communities most impacted by these disparities. Such policies should
prioritize investments in healthcare infrastructure, culturally competent care, and
programs that support at-risk populations.

## Addressing Public Health and Long-Term Implications of Restrictive Abortion Laws

Addressing Public Health and Long-Term Implications of Restrictive Abortion Laws

Restrictive abortion policies significantly impact maternal and infant health, as
evidenced by an increase of 0.285 infant deaths per 1,000 live births in states enforcing
abortion bans during the post-injunction period. These policies exacerbate systemic issues
such as inadequate prenatal care, gaps in healthcare access, and socioeconomic inequities,
disproportionately affecting marginalized communities. Over time, the consequences may
include higher rates of untreated maternal mental health issues, delayed or insufficient
prenatal care, and elevated infant mortality, creating lasting health challenges for
families and communities. Addressing these disparities requires targeted interventions,
such as expanding maternal mental health services, improving prenatal and postnatal care
in underserved areas, ensuring culturally competent healthcare, and advocating for
equitable policies. Investments in community-based support systems, enhanced healthcare
delivery, and robust data infrastructure are essential to monitor and mitigate the
long-term impacts of these policies, fostering sustainable improvements in maternal and
infant health outcomes.

## Data and Temporal Limitations

A key limitation of our analysis is the temporal and demographic scope of the dataset. The
data includes only infant deaths under 1 year of age occurring within the United States to
U.S. residents during 2021–2022. This timeframe does not capture the full year of infant
mortality outcomes following the Dobbs decision, as infants who died in 2023 are not
included. Furthermore, the dataset’s focus on infants under 1 year of age may exclude
cases that reflect longer-term health impacts. These limitations could lead to an
underestimation of the full effects of abortion bans. As no publicly available dataset
currently provides complete 2023 data or longer-term health outcomes, future research
should incorporate more detailed and updated datasets to provide a clearer and more
complete analysis of the long-term trends and implications of restrictive abortion
policies.

## Weaknesses and Future Directions

While this study provides meaningful findings, several limitations should be addressed in
future research. The analysis relies on aggregate state-level data, which may obscure
localized effects and intra-state disparities. Unmeasured confounders, such as economic
conditions, healthcare infrastructure differences, or state-specific policies, could also
influence the results. Moreover, the short timeframe of the dataset limits the ability to
assess the long-term effects of abortion policies. Future research should integrate more
granular data, such as county-level analyses, and extend the study period to capture
longer-term impacts. Qualitative studies focusing on the lived experiences of affected
populations can complement quantitative analyses, offering a deeper understanding of the
broader implications of restrictive abortion laws. These efforts will support the
development of evidence-based policies aimed at reducing disparities and improving
maternal and infant health outcomes.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details {#sec-data-details}

## Dataset and Graph Sketches

Sketches depicting both the desired dataset and the graphs generated in this analysis is
available in the GitHub Repository `other/sketches`.

## Data Cleaning

The raw dataset, consisting of U.S. infant mortality records from 2021 and 2022, was
cleaned and pre-processed to ensure its suitability for analysis. The dataset was first
imported from a tab-separated file, with irregular lines handled appropriately. Variables
were standardized using `janitor::clean_names()` to ensure consistent naming conventions.
Key variables such as state, maternal age, maternal race, year of death, month, and infant
death rates were selected for analysis, and missing values were removed.

Additional variables were created to facilitate analysis: `after_injunction`, indicating
whether the data correspond to the period after June 2022 (post-Dobbs decision), and
`abortion_illegal`, denoting whether the state had restrictive abortion laws. Death rate
reliability was assessed, and unreliable rates were flagged for transparency. The death
rate column was cleaned to remove annotations and converted to numeric format for
analysis. A date variable was created by combining the year and month columns, providing a
timeline for analysis. Finally, a date variable was created by combining the year and
month variables. The cleaned dataset was saved in CSV and Parquet formats to ensure
compatibility with analytical workflows.

# Model details {#sec-model-details}

## Diagnostics

\newpage

# References
